---
title: "How to follow Insper's design guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to follow Insper's design guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

Insper's brand guide covers print, not charts. It specifies Pantone
references, tints and shades per hue family, and rules for text on colored
backgrounds, but no categorical sequence and no guidance on which palette
suits which kind of data. This vignette records how `insperplot` fills those
gaps, which of its palettes are brand tokens verbatim, and which are derived.

```{r setup}
library(insperplot)
library(ggplot2)
```

## Accessibility

The `main` palette is the brand's own hue set, and it puts red, green, orange and yellow next to one another. That combination is hard to read under the common forms of color vision deficiency. Deuteranopia and protanopia both compress the red-green axis, so categories that look distinct to you may merge for roughly 1 in 12 male readers.

```{r}
insper_palette("main")
```

Use the `colorblind` palette when telling categories apart matters more than brand fidelity: a chart with many categories, a slide seen from the back of a room, or a document whose rendering you cannot control. It is the [Okabe-Ito](https://jfly.uni-koeln.de/color/) set, designed so that its colors stay distinct under the usual color vision deficiencies.

```{r}
insper_palette("colorblind")
```

It is the only palette in `insperplot` with no Insper colors. Use `main` for on-brand work, and switch to `colorblind` when you have a reason to.

A separate question is what color text should sit *on top of* a brand color. The brand guide rules on this for the family base colors and neutrals, and the `acessibilidade` column of the `insper_color_reference` dataset records those rules.

```{r}
subset(
  insper_color_reference,
  !is.na(acessibilidade),
  select = c(nome, hex, acessibilidade)
)
```

The guide does not rule on the tints and shades, so those rows are `NA`. For labels drawn on colored shapes, `insperplot` computes WCAG relative luminance at draw time instead. That is how `insper_barplot()` picks a white or dark label for each segment of stacked and filled bars.

## Typography

### Brand guidelines

Insper's brand guidelines assign three fonts.

- GT Ultra Fine Bold carries titles, subtitles and most highlights.
- Acumin Pro ExtraCondensed Bold carries other highlights, always in uppercase.
- Inter carries body text, with Verdana as a fallback.

Following them literally gives the code below.

```{r, eval=FALSE}
theme_minimal(base_family = "Inter 18pt") +
  theme_sub_plot(
    title = element_text(family = "GT Ultra Fine Bold", size = 22),
    subtitle = element_text(family = "GT Ultra Fine Bold", size = 10),
    caption = element_text(family = "Inter 18pt")
  )

some_plot +
  annotate(
    "text",
    label = "SOME HIGHLIGHT ALWAYS IN CAPS",
    family = "Acumin Pro ExtraCondensed"
  )
```

This scheme is hard to reproduce. GT Ultra is a proprietary, paid font. Acumin Pro and Verdana ship with some operating systems but not others. Inter ships with few operating systems, but it is free to download from Google Fonts.

### `insperplot` implementation

`insperplot` bundles Inter, so the font installs with the package. Inter carries all text except the title, which uses Georgia, a serif font that ships with most operating systems.

```{r, eval=FALSE}
tinsper <- theme_insper()

tinsper$plot.title@family
#> "Georgia"

tinsper$plot.subtitle@family
#> "Inter"

tinsper$text@family
#> "Inter"
```

The default `theme_insper()` produces a plot like the one below.

```{r}
base_plot <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(
    title = "Basic Theme",
    subtitle = "USING THE PACKAGE TO MAKE PLOTS",
    x = "Miles per Gallon",
    y = "Weight (lbs)",
    caption = "Source: mtcars | Insper"
  )

base_plot +
  theme_insper()
```

Users who have GT Ultra, Acumin, or Verdana installed can use them by setting `font_title` and `font_text` in `theme_insper()`.

```{r, eval = FALSE}
base_plot +
  theme_insper(font_title = "GT Ultra", font_text = "Verdana")
```

Where Acumin is installed, it can carry subtitles and highlights. The chunks below are not run, because they name fonts that the package does not bundle.

```{r, eval = FALSE}
base_plot +
  theme_insper(
    font_title = "Georgia",
    font_text = "Inter"
  ) +
  theme_sub_plot(
    subtitle = element_text(family = "Acumin Pro ExtraCondensed", size = 14)
  )
```

Verdana is the brand's fallback for Inter. Since Inter ships with the package, you rarely need it.

```{r, eval = FALSE}
base_plot +
  theme_insper(
    font_title = "Georgia",
    font_text = "Verdana"
  )
```

## Colors

Insper's brand guidelines target printed materials, not charts. The kit specifies Pantone references and a handful of tints and shades per hue family, but no categorical sequence. The bundled palettes will not suit every chart, so it helps to know how each one was built.

### Where each palette comes from

Some palettes are the brand kit verbatim.

- `main` is the six secondary hue bases plus vermelho, interleaved so that adjacent categories contrast.
- The six hue sequentials (`turquesa`, `verde`, `amarelo`, `laranja`, `rosa`, `roxo`) are the kit's own five tokens per family, sorted light to dark by perceptual lightness. No color was invented.
- `cinza` is the five neutral grays.

Others are derived, because the kit does not ship what a chart needs.

- `muted` is `main` desaturated by 25%. None of its seven hexes appear in the guide. It exists because full-chroma brand hues are overwhelming across large filled areas.
- `vermelho` and `azul` are interpolated ramps. The kit ships only a single base token for each, so the other steps were interpolated in CIELAB. Only the base token is official: it sits at the middle of `vermelho` and at the dark end of `azul`.
- The four diverging palettes have brand colors at both arms but share a neutral midpoint, `#F4F4F2`, which is not a brand token.

Finally, `colorblind` is the Okabe-Ito palette, which contains no Insper colors.

### Building a categorical set by hand

When `main` is too loud and `muted` is not distinct enough, you can pull mid-range steps out of the sequential ramps. Those steps are still brand colors, and finding a set that works is mostly trial and error.

```{r}
paleta <- c(
  insper_palette("turquesa")[4],
  insper_palette("amarelo")[3],
  insper_palette("roxo")[2],
  insper_palette("muted")[1],
  insper_palette("cinza")[3],
  insper_palette("verde")[3],
  insper_palette("laranja")[3]
)
```

The indices were chosen by eye: darker steps for the light families (turquesa, amarelo, verde, laranja) and a lighter step for roxo, which is very dark at its base. The result is shown below on a stacked bar chart and a stacked area chart.


```{r, echo = FALSE}
dat_barplot <- data.frame(
  x = rep(c("A", "B", "C"), each = 5),
  group = rep(paste0("Group", 1:5), 3),
  y = c(40, 20, 20, 10, 10, 20, 35, 15, 15, 15, 10, 10, 20, 30, 30)
)

ggplot(dat_barplot, aes(x, y, fill = group)) +
  geom_col() +
  scale_fill_manual(values = paleta) +
  guides(fill = "none") +
  theme_insper()

recent_fuel <- subset(fossil_fuel, year >= 1960)

ggplot(recent_fuel, aes(year, consumption, fill = fuel)) +
  geom_area() +
  scale_fill_manual(values = paleta) +
  scale_x_continuous(expand = expansion(mult = c(0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_insper(border = "half")
```

## Insper Cidades

The Centro de Estudos das Cidades (Laboratório Arq.Futuro) publishes its own application manual. It adds four colors to the shared institutional vermelho, branco and preto.

```{r}
insper_palette("cidades")
```

These are `folhagem`, `solar`, `asfalto` and `tijolo`. Everything prefixed `cidades` in `insperplot` comes from that manual rather than the institutional brand kit, and nothing in the package defaults to them.

### The sequential ramps are derived

Unlike the six institutional hue families, the Cidades manual ships no tints or shades, only the four base colors. The package therefore builds each ramp: seven steps interpolated in CIELAB from a lightened anchor, through the official color, to a darkened one. The official color is pinned at position 4.

```{r}
insper_palette("cidades_folhagem")
```

Without the pin, rounding in the interpolation would shift the official color slightly: `folhagem` would come back from its own ramp as `#009880` rather than `#019881`.

### Text contrast

The Cidades manual has no accessibility page, so unlike the institutional colors there is no official ruling on what text color to put on these backgrounds. The values below are computed from WCAG relative luminance rather than taken from a guide.

| Color | Contrast vs white | Contrast vs black | Use |
| --- | --- | --- | --- |
| `folhagem` | 3.61 | 5.81 | black text |
| `solar` | 1.95 | 10.78 | black text |
| `asfalto` | 9.01 | 2.33 | white text |
| `tijolo` | 3.36 | 6.26 | black text |

Asfalto is the only one of the four that takes white text.

### Diverging data

There are two Cidades diverging palettes.

```{r}
insper_palette("cidades_folhagem_asfalto")
insper_palette("cidades_folhagem_tijolo")
```

Both share the same neutral midpoint (`#F4F4F2`) as the institutional diverging palettes, so they drop into `scale_fill_insper_c()` the same way.
