---
title: "How to follow Insper's design guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to follow Insper's design guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

# Introduction

While Insper has very specific brand guidelines, it lacks a data visualization module.

This vignette is a small step towards filling this gap.

## Basics

```{r setup}
library(insperplot)
library(ggplot2)
```

### Accessibility

The `main` palette is the brand's own hue set, and it puts red, green, orange and yellow next to one another. That is a difficult combination under the common forms of color vision deficiency: deuteranopia and protanopia both compress the red-green axis, so categories that look obviously distinct to you may collapse into one another for roughly 1 in 12 male readers.

```{r}
insper_palette("main")
```

When distinguishability matters more than brand fidelity — a chart with many categories, a slide seen from the back of a room, a document you cannot control the rendering of — use the `colorblind` palette instead. It is the [Okabe-Ito](https://jfly.uni-koeln.de/color/) set, chosen precisely so that its colors stay separable across the usual color vision deficiencies.

```{r}
insper_palette("colorblind")
```

This is the one palette in `insperplot` that contains no Insper colors, and that is deliberate. Treat it as an escape hatch rather than a default: reach for `main` for on-brand work, and switch when you have a reason to.

A second, separate question is what color text should sit *on top of* a brand color. The brand guide rules on this directly for the family base colors and neutrals, and those rules are shipped in the `acessibilidade` column of the `insper_color_reference` dataset:

```{r}
subset(
  insper_color_reference,
  !is.na(acessibilidade),
  select = c(nome, hex, acessibilidade)
)
```

The guide does not rule on the tints and shades, so those rows are `NA`. For those, `insperplot` falls back to computing WCAG relative luminance at draw time — this is what picks the per-bar label color in `insper_barplot()` when you use the stacked or filled variants, so labels stay readable without you specifying anything.

## Typography

### Brand guidelines

According to Insper's brand guidelines:

- GT Ultra Fine Bold should be used in titles, subtitles and most highlights.
- Acumin Pro ExtraCondensed Bold should be used in other highlights always in uppercase.
- Inter should be used in body text with Verdana as a fallback.

This results in something like the code below.

```{r, eval=FALSE}
theme_minimal(base_family = "Inter 18pt") +
  theme_sub_plot(
    title = element_text(family = "GT Ultra Fine Bold", size = 22),
    subtitle = element_text(family = "GT Ultra Fine Bold", size = 10),
    caption = element_text(family = "Inter 18pt")
  )

some_plot +
  annotate(
    "text",
    label = "SOME HIGHLIGHT ALWAYS IN CAPS",
    family = "Acumin Pro ExtraCondensed",
  )
```

There are several problems with this typography scheme. GT Ultra is a proprietary (and expensive) font, Acumin Pro and Verdana are available on some operating systems but not on others. Inter is not available on most operating systems, but is freely available for download from Google Fonts.

### `insperplot` implementation

The approach used by `insperplot` tries to be safe: it bundles Inter, meaning the font is installed together with the package, and uses safe fallbacks for titles. Essentially, the package uses Inter for everything except for the title, which uses `'Georgia'`.

```{r, eval=FALSE}
tinsper <- theme_insper()

tinsper$plot.title@family
#> "Georgia"

tinsper$plot.subtitle@family
#> "Inter"

tinsper$text@family
#> "Inter"
```

The default `theme_insper()` will produce a plot like the one below.
```{r}
base_plot <- ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(
    title = "Basic Theme",
    subtitle = "USING THE PACKAGE TO MAKE PLOTS",
    x = "Miles per Gallon",
    y = "Weight (lbs)",
    caption = "Source: mtcars | Insper"
  )

base_plot +
  theme_insper()
```

Users that have GT Ultra, Acumin, or Verdana installed can use them by setting `font_title` and `font_text` in `theme_insper()`.

```{r, eval = FALSE}
base_plot +
  theme_insper(font_title = "GT Ultra", font_text = "Verdana")
```

Where Acumin is installed, it can carry subtitles and highlights. Neither of the two chunks below is evaluated when this vignette is built, because the fonts they name are not bundled and `element_text()` falls back silently when a family is missing, which would show you a rendered figure in the wrong font.

```{r, eval = FALSE}
base_plot +
  theme_insper(
    font_title = "Georgia",
    font_text = "Inter"
  ) +
  theme_sub_plot(
    subtitle = element_text(family = "Acumin Pro ExtraCondensed", size = 14)
  )
```

Verdana ships with most operating systems, but it buys little here, since Inter is bundled with the package.

```{r, eval = FALSE}
base_plot +
  theme_insper(
    font_title = "Georgia",
    font_text = "Verdana"
  )
```

## Colors

Insper's brand guidelines are written for printed materials, with little emphasis on data visualization. The kit specifies Pantone references and a handful of tints and shades per hue family; it does not specify a categorical sequence. As a result, the bundled palettes will not always work out of the box, and it helps to know how each one was built.

### Where each palette comes from

Some palettes are the brand kit verbatim:

- `main` is the six secondary hue bases plus vermelho, interleaved so that adjacent categories contrast.
- The six hue sequentials — `turquesa`, `verde`, `amarelo`, `laranja`, `rosa`, `roxo` — are the kit's own five tokens per family, sorted light to dark by perceptual lightness. No color was invented.
- `cinza` is the five neutral grays.

Others are derived, because the kit does not ship what a chart needs:

- `muted` is `main` desaturated by 25%. None of its seven hexes appear in the guide. It exists because full-chroma brand hues are overwhelming across large filled areas.
- `vermelho` and `azul` are interpolated ramps. The kit ships no tints or shades for either, only the single base token, so the intermediate steps were generated perceptually and only the endpoints are official.
- The four diverging palettes have brand colors at both arms but share a neutral midpoint, `#F4F4F2`, which is not a brand token.

And `colorblind` is derived from the Okabe-Ito palette, which is not an Insper palette.

### Building a categorical set by hand

When `main` is too loud and `muted` is not distinct enough, you can pull mid-range steps out of the sequential ramps. Those steps are still brand colors, and getting them to work is mostly a trial and error process.

```{r}
paleta <- c(
  insper_palette("turquesa")[4],
  insper_palette("amarelo")[3],
  insper_palette("roxo")[2],
  insper_palette("muted")[1],
  insper_palette("cinza")[3],
  insper_palette("verde")[3],
  insper_palette("laranja")[3]
)
```

The indices are chosen by eye rather than by rule: darker steps for the light families (turquesa, amarelo, verde, laranja) and a lighter step for roxo, which is very dark at its base. The result is shown below on a stacked bar chart and a stacked area chart.


```{r, echo = FALSE}
dat_barplot <- data.frame(
  x = rep(c("A", "B", "C"), each = 5),
  group = rep(paste0("Group", 1:5), 3),
  y = c(40, 20, 20, 10, 10, 20, 35, 15, 15, 15, 10, 10, 20, 30, 30)
)

ggplot(dat_barplot, aes(x, y, fill = group)) +
  geom_col() +
  scale_fill_manual(values = paleta) +
  guides(fill = "none") +
  theme_insper()

recent_fuel <- subset(fossil_fuel, year >= 1960)

ggplot(recent_fuel, aes(year, consumption, fill = fuel)) +
  geom_area() +
  scale_fill_manual(values = paleta) +
  scale_x_continuous(expand = expansion(mult = c(0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_insper(border = "half")
```

## Insper Cidades

The Centro de Estudos das Cidades — Laboratório Arq.Futuro publishes its own application manual, with four colors of its own alongside the shared institutional vermelho, branco and preto:

```{r}
insper_palette("cidades")
```

These are `folhagem`, `solar`, `asfalto` and `tijolo`. Everything prefixed `cidades` in `insperplot` comes from that manual rather than the institutional brand kit, and nothing in the package defaults to them — they are opt-in.

### The sequential ramps are derived

Unlike the six institutional hue families, the Cidades manual ships no tints or shades — just the four standalone colors. The ramps are therefore built here, not transcribed: seven steps interpolated in CIELAB from a lightened anchor through the official color to a darkened one, with the official color pinned at position 4.

```{r}
insper_palette("cidades_folhagem")
```

The pin matters more than it sounds. Lab interpolation rounds endpoints by a bit, so without it `folhagem` would come back from its own ramp as `#009880` rather than the official `#019881`.

### Text contrast

The Cidades manual has no accessibility page, so unlike the institutional colors there is no official ruling on what text color to put on these backgrounds. The values below are computed from WCAG relative luminance rather than taken from a guide:

| Color | Contrast vs white | Contrast vs black | Use |
| --- | --- | --- | --- |
| `folhagem` | 3.61 | 5.81 | black text |
| `solar` | 1.95 | 10.78 | black text |
| `asfalto` | 9.01 | 2.33 | white text |
| `tijolo` | 3.36 | 6.26 | black text |

Asfalto is the only one of the four that takes white text.

### Diverging data

There are two Cidades diverging palettes.

```{r}
insper_palette("cidades_folhagem_asfalto")
insper_palette("cidades_folhagem_tijolo")
```

Both share the same neutral midpoint (`#F4F4F2`) as the institutional diverging palettes, so they drop into `scale_fill_insper_c()` the same way.
