---
title: "Getting Started with insperplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with insperplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  fig.width = 7,
  fig.asp = 0.618,
  fig.align = "center",
  warning = FALSE,
  message = FALSE,
  eval = rlang::is_installed(c("dplyr", "lubridate", "RcppRoll"))
)
```

**insperplot** extends ggplot2 with Insper's visual identity, providing custom themes, color palettes, and specialized plotting functions for academic and institutional use. This vignette walks through the theme, the color scales, and the convenience plot functions.

## Installation

Install the latest release from the Insper Cidades [r-universe](https://inspercidades.r-universe.dev), which ships pre-built binaries for Windows, macOS, and Linux.

```{r eval=FALSE}
install.packages(
  "insperplot",
  repos = c(
    "https://inspercidades.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

The development version installs from GitHub.

```{r eval=FALSE}
remotes::install_github("inspercidades/insperplot")
```

We recommend using `insperplot` alongside the `ggplot2` package.
```{r setup}
library(insperplot)
library(ggplot2)
library(dplyr)
```

## Basic Usage: theme_insper()

`theme_insper()` applies Insper's typography and colors to a ggplot2 plot. Add it as a layer like any other theme.

```{r theme-basic}
ggplot(mtcars, aes(mpg, wt)) +
  geom_point() +
  labs(
    title = "Basic Theme",
    subtitle = "Using theme_insper()",
    caption = "Source: mtcars"
  ) +
  theme_insper()
```

### Customizing the Theme

`theme_insper()` supports several arguments for quick customization.

- `font_title`: Font for titles. Default is `"Georgia"`.
- `font_text`: Font for body text. Default is `"Inter"`.
- `grid`: Grid lines. Default is `TRUE`.
- `border`: Border style. Options are `"none"` (default), `"half"`, or `"closed"`.
- `align`: Title, subtitle, and caption alignment. Default is `"panel"`.

The following examples use `macro_series`, a dataset bundled with insperplot. It contains monthly Brazilian macroeconomic indicators — including inflation (IPCA), industrial production (IPI), and economic activity (IBC-Br) — sourced from the Brazilian Central Bank. See `?macro_series` for details.

Here we plot the IBC-Br index (a monthly GDP proxy) with a `"half"` border and a sans-serif title font.

```{r theme-border-half}
ggplot(macro_series, aes(date, ibcbr_dessaz)) +
  geom_line(lwd = 0.7) +
  labs(
    title = "Economic Activity Index",
    subtitle = "IBC-Br, seasonally adjusted (base 2002 = 100)",
    x = NULL,
    y = "IBC-Br",
    caption = "Source: Banco Central do Brasil"
  ) +
  theme_insper(font_title = "Inter", border = "half")
```

### Fine-Tuning with theme_sub_*()

For more granular control, ggplot2 4.0+ provides `theme_sub_*()` helper functions that target specific theme components. Add them after `theme_insper()` to override individual elements without resetting the whole theme.

In the example below, we add axis ticks and lines to the x-axis with `theme_sub_axis_x()`, and change the panel background with `theme_sub_panel()`.

```{r theme-sub-override}
ggplot(macro_series, aes(date, ibcbr_dessaz)) +
  geom_line(lwd = 0.7) +
  labs(
    title = "Economic Activity Index",
    x = NULL,
    y = "IBC-Br",
    caption = "Source: Banco Central do Brasil"
  ) +
  theme_insper() +
  theme_sub_axis_x(
    ticks = element_line(color = "gray20", linewidth = 0.25),
    line = element_line(color = "gray20", linewidth = 0.5)
  ) +
  theme_sub_panel(
    background = element_rect(fill = "#f3f3f3", color = NA)
  )
```

## Working with Colors

Use `insper_palette()` to preview a specific palette and `show_insper_palettes()` to see all available palettes at a glance.

```{r palettes-overview, fig.height=4}
insper_palette("vermelho")
```

```{r palettes-all, fig.height=7}
show_insper_palettes()
```

Palettes come in three types — qualitative for categorical data, sequential for ordered magnitudes, and diverging for data with a meaningful centre — and in three families.

- **Institutional** (`main`, `muted`, and the single-hue ramps like `vermelho` and `turquesa`) is built from Insper's 2026 brand kit and is what every function defaults to.
- **Insper Cidades** (everything prefixed `cidades`) belongs to the Centro de Estudos das Cidades sub-brand. Use it for that centre's work, and don't mix it with the institutional palettes in one chart — the two families are close enough in color that a chart using both looks like a rendering error.
- **`colorblind`** is the Okabe-Ito palette. It contains no Insper colors and exists as an accessibility fallback for when the brand hues can't be told apart by all readers.

`show_insper_palettes()` also returns a data frame of palette metadata invisibly, including a `recommended_use` column, and accepts a `type` filter.

```{r palettes-filter, eval=FALSE}
show_insper_palettes("diverging")

meta <- show_insper_palettes()
subset(meta, type == "qualitative")
```

The design guide vignette, `vignette("brand-kit", package = "insperplot")`, covers palette provenance and accessibility in more depth.

### Using Colors in ggplot2

The package provides `scale_color_insper_*()` and `scale_fill_insper_*()` functions for integrating Insper colors into ggplot2 plots. Use the `_d` variants for discrete (categorical) data and `_c` for continuous (numeric) data.

#### Discrete Scales

For categorical variables, use `scale_color_insper_d()` or `scale_fill_insper_d()`.

The examples below use `fossil_fuel`, a dataset bundled with insperplot. It tracks global primary energy consumption (in TWh) by fuel type (coal, oil, and gas) from 1800 to the present, sourced from [Our World in Data](https://ourworldindata.org/grapher/global-fossil-fuel-consumption). See `?fossil_fuel` for details.

First, a stacked area chart of consumption since 1960 using `scale_fill_insper_d()`.

```{r scale-discrete-fill}
recent_fuel <- subset(fossil_fuel, year >= 1960)

ggplot(recent_fuel, aes(year, consumption, fill = fuel)) +
  geom_area() +
  scale_fill_insper_d(palette = "main") +
  scale_x_continuous(expand = expansion(mult = c(0))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  theme_insper(border = "half")
```

The same data shown as a line chart with `scale_color_insper_d()`.

```{r scale-discrete-color}
ggplot(fossil_fuel, aes(year, consumption, color = fuel)) +
  geom_line() +
  scale_color_insper_d(palette = "main") +
  theme_insper() +
  labs(
    title = "Global Fossil Fuel Consumption",
    x = NULL,
    y = "Consumption (TWh)",
    color = "Fuel",
    caption = "Source: Our World in Data (CC BY 4.0)"
  )
```

#### Continuous Scales

For numeric variables, use `scale_color_insper_c()` or `scale_fill_insper_c()`. These create smooth color gradients from any Insper palette.

The heatmap below shows 12-month rolling inflation (IPCA) across years and months using the `"vermelho"` palette. We compute the rolling product from the `macro_series` dataset using `lubridate` (for date parts) and `RcppRoll` (for the rolling window).

```{r scale-continuous-fill}
#| fig-width: 8
#| fig-height: 9
macro_calendar <- macro_series |>
  mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    ipca12m = RcppRoll::roll_prodr(1 + ipca / 100, n = 12) - 1,
    ipca12m = ipca12m * 100
  ) |>
  filter(between(year, 2004, 2024))

ggplot(macro_calendar, aes(month, year, fill = ipca12m)) +
  geom_tile(color = "white", lwd = 0.1) +
  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb,
    position = "top",
    expand = expansion(c(0))
  ) +
  scale_y_continuous(breaks = 2004:2024, expand = expansion(c(0))) +
  scale_fill_insper_c(
    name = "IPCA (YoY, %)",
    palette = "vermelho"
  ) +
  labs(x = NULL, y = NULL) +
  theme_insper() +
  theme_sub_legend(
    title = element_text(family = "Georgia", size = 12),
    title.position = "top",
    key.width = unit(4, "lines"),
    position = "top"
  )
```

## Convenience Plot Functions

Beyond the theme and scales, insperplot provides high-level plotting functions that combine common geom + scale + theme patterns into a single call. Each function returns a standard ggplot object, so you can still add layers, labs, or further customization with `+`.

### Time Series

`insper_timeseries()` creates line plots optimized for temporal data. It handles `Date` and `POSIXct` axes automatically and supports multiple series via the `color` argument.

```{r plot-timeseries}
recent_macro <- subset(macro_series, date >= as.Date("2015-01-01"))

insper_timeseries(recent_macro, x = date, y = ipca) +
  labs(
    title = "Brazilian Inflation (IPCA)",
    subtitle = "Monthly rate, 2015–present",
    x = NULL,
    y = "IPCA (%)",
    caption = "Source: Banco Central do Brasil"
  )
```

## Saving Plots

For best results when saving plots, use `ggsave()` with the [ragg](https://ragg.r-lib.org/) device to ensure proper font rendering.

```{r save-ggsave, eval=FALSE}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() +
  theme_insper()

ggsave(
  "my_plot.png",
  p,
  width = 8,
  height = 5,
  dpi = 300,
  device = ragg::agg_png
)
```

Alternatively, `save_insper_plot()` wraps `ggsave()` with sensible defaults for Insper-branded output: dimensions in centimeters, a golden-ratio aspect ratio, and automatic ragg device selection for PNG files when the package is installed.

```{r save-insper, eval=FALSE}
save_insper_plot(p, "my_plot.png")

# Custom dimensions (in cm by default)
save_insper_plot(p, "my_plot.png", width = 16, height = 10)
```

## Available Functions

In addition to `insper_timeseries()` shown above, insperplot includes convenience functions for other common chart types.

| Function | Chart type |
|---|---|
| `insper_timeseries()` | Line plots for temporal data |
| `insper_barplot()` | Vertical and horizontal bar charts |
| `insper_scatterplot()` | Scatter plots |
| `insper_area()` | Area charts |
| `insper_boxplot()` | Box plots |
| `insper_violin()` | Violin plots |
| `insper_histogram()` | Histograms |
| `insper_density()` | Density plots |
| `insper_heatmap()` | Heatmaps |

All functions return ggplot objects, apply `theme_insper()` by default, and accept a `palette` argument for Insper color palettes. See their help pages (e.g. `?insper_barplot`) for full documentation.

## Learn More

- [Package website](https://inspercidades.github.io/insperplot/) — full function reference and additional articles
- [GitHub repository](https://github.com/inspercidades/insperplot) — source code, issues, and contributing
- [Insper brand colors](https://github.com/Insper/Cores-Marca-Insper) — the color system this package is built on
