---
title: "Color Reference Table"
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

This page is a searchable quick-reference for every official Insper brand color: Pantone spot color, RGB, and hex values, organized by hierarchy (Principais, Secundários, Neutros) and hue family. It is generated directly from the `insper_color_reference` dataset shipped with the package.

For programmatic access in plots, use `insper_palette()` or the `scale_*_insper_*()` functions instead — this table is documentation, not an API.

```{r setup}
library(insperplot)
library(reactable)
library(htmltools)
```

## Table

```{r color-table, echo = FALSE}
swatch_cell <- function(hex) {
  div(
    style = paste0(
      "background-color:",
      hex,
      ";",
      "border-radius:4px;",
      "border:1px solid rgba(0,0,0,0.1);",
      "height:1.4rem;width:100%;"
    ),
    title = hex
  )
}

tab <- dplyr::select(insper_color_reference, -familia)

reactable(
  tab,
  # groupBy = "hierarquia",
  defaultPageSize = 50,
  filterable = TRUE,
  searchable = TRUE,
  highlight = TRUE,
  striped = TRUE,
  columns = list(
    hierarquia = colDef(name = "Hierarquia"),
    familia = colDef(name = "Família"),
    nome = colDef(name = "Nome"),
    amostra = colDef(
      name = "Amostra",
      cell = function(value) swatch_cell(value),
      filterable = FALSE,
      width = 90
    ),
    pantone = colDef(name = "Pantone"),
    rgb = colDef(name = "RGB"),
    hex = colDef(name = "Hex"),
    acessibilidade = colDef(
      name = "Texto",
      cell = function(value, index) {
        if (is.na(value)) {
          return("")
        }
        bg <- tab$hex[index]
        label <- if (value == "ambos") "Ambos" else "Insper"
        font <- if (value == "branco") "#FFFFFF" else "#000000"
        div(
          style = paste0(
            "background-color:", bg, ";",
            "color:", font, ";",
            "border-radius:4px;",
            "text-align:center;",
            "font-weight:600;",
            "padding:0.1rem 0.4rem;"
          ),
          label
        )
      },
      filterable = FALSE,
      width = 100
    )
  ),
  defaultColDef = colDef(headerStyle = list(fontWeight = 600))
)
```

## Acessibilidade: text on colored backgrounds

The brand guide rules on which text (or logo) color may sit on each base background: white only, black only, or both. The `acessibilidade` column above condenses those rules — the swatch in the "Texto" column shows the permitted color in use. Tints and shades (`_1`–`_4`) are not covered by the guide, so they are blank.

```{r accessibility-guide, echo = FALSE, fig.width = 8, fig.height = 3.5, fig.alt = "Grid of Insper base colors, each swatch labeled with sample text in the permitted color: white, black, or both."}
library(ggplot2)

acc <- subset(insper_color_reference, !is.na(acessibilidade))
acc$col <- (seq_len(nrow(acc)) - 1) %% 5
acc$row <- ((seq_len(nrow(acc)) - 1) %/% 5) * 1.4

labels <- do.call(
  rbind,
  lapply(seq_len(nrow(acc)), function(i) {
    r <- acc[i, ]
    text_cols <- switch(
      r$acessibilidade,
      branco = "#FFFFFF",
      preto = "#000000",
      ambos = c("#FFFFFF", "#000000")
    )
    offsets <- if (length(text_cols) == 2) c(-0.16, 0.16) else 0
    data.frame(
      col = r$col,
      row = r$row,
      y = r$row + offsets,
      text_color = text_cols,
      hex = r$hex,
      nome = r$nome
    )
  })
)

ggplot(acc, aes(x = col, y = row)) +
  geom_tile(aes(fill = hex), width = 0.92, height = 0.85, color = "grey85", linewidth = 0.3) +
  geom_text(
    data = labels,
    aes(y = y, label = "Insper", color = text_color),
    fontface = "bold",
    size = 3.6
  ) +
  geom_text(
    aes(y = row - 0.55, label = nome),
    size = 3,
    color = "grey30"
  ) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_y_reverse() +
  coord_equal(ratio = 0.8) +
  theme_void()
```

The same logic is applied automatically inside the package: plot functions that draw labels on colored areas (e.g. `insper_barplot()` with `text = TRUE`) pick white or dark text per segment based on the fill's luminance.

## Source

Values are transcribed from Insper's 2026 brand guide via `data-raw/colors_palettes.R` and assembled into this flat table by `data-raw/create_color_reference_table.R`. See `?insper_color_reference` for column definitions. A companion `data-raw/insper_color_reference.xlsx` file mirrors this table with color-filled swatch cells for use outside R.
